---
title: "Atlas integration"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Setup chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/10-atlas-integration.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align = "center",
  out.width = '100%'
)
```

Load Bioconductor packages:

```{r}
library(batchelor)
library(EnsDb.Mmusculus.v79)
library(MouseGastrulationData)
library(org.Mm.eg.db)
library(pheatmap)
library(scater)
library(scran)
```

Load CRAN packages:

```{r}
library(msigdbr)
```

Source user-defined functions:

```{r}
source("code/cleanSCE.R")
source("code/cleanEmbryoAtlasData.R")
source("code/splitByCol.R")
```

## Data

### Query data

Read experiment object:

```{r cache = FALSE}
sce <- readRDS("output/09-cell-annotation.rds")
```

Clean experiment object:

```{r cache = FALSE}
sce <- cleanSCE(sce, sample = "AL_KDR_HET_C")
```

### Reference data

Read experiment object:

```{r cache = FALSE}
sel <- subset(AtlasSampleMetadata, stage == "E8.25")

ref <- EmbryoAtlasData(type = "processed", samples = sel$sample)
```

Clean experiment object:

```{r cache = FALSE}
ref <- cleanEmbryoAtlasData(ref)
```

Split experiment object by sample:

```{r cache = FALSE}
ref <- splitByCol(ref, ref$sample)
```

## Integration

### Data processing

Create list of experiment data:

```{r cache = FALSE}
all <- c(ref, "AL_KDR_HET_C" = sce)
```

```{r include = FALSE}
rm(ref, sce); gc() # delete large objects and trigger garbage collection
```

Subset data to common features:

```{r cache = FALSE}
ids <- Reduce(intersect, lapply(all, rownames))

all <- lapply(all, "[", i = ids)
```

### Variance modelling

Model the per-gene variance:

```{r cache = FALSE}
dec <- lapply(all, modelGeneVar)
```

Combine variance decompositions:

```{r}
dec <- do.call(combineVar, dec)

rownames(dec) <- Reduce(intersect, lapply(all, rownames))
```

### Feature selection

Identify features specified to be on a blacklist:

```{r}
ids <- read.csv("data/blacklist.csv")

dec$blacklist <- rownames(dec) %in% ids$gene_id
```

Identify highly variable genes:

```{r}
hvg <- subset(dec, blacklist == FALSE)

hvg <- getTopHVGs(hvg, var.field = "bio", var.threshold = 0, fdr.field = "p.value", fdr.threshold = 0.05)
```

### Batch normalization

Perform scaling normalization:

```{r cache = FALSE}
all <- do.call(multiBatchNorm, all)
```

Combine samples into experiment object:

```{r cache = FALSE}
all <- do.call(cbind, all)
```

### Batch integration

Create sequential merge order:

```{r}
dat <- subset(AtlasSampleMetadata, stage %in% all$stage)

lvl <- c("E6.5", "E6.75", "E7.0", "mixed_gastrulation", "E7.25", "E7.5", "E7.75", "E8.0", "E8.25", "E8.5")

dat$stage <- factor(dat$stage, levels = intersect(lvl, all$stage))

dat <- dat[order(dat$stage, dat$ncells, decreasing = TRUE), ]
  
idx <- c(as.character(dat$sample), "AL_KDR_HET_C")
```

Correct experiments using nearest neighbours:

```{r cache = FALSE}
set.seed(163052521)

mnn <- correctExperiments(
  all,
  batch = all$sample,
  subset.row = hvg,
  correct.all = TRUE,
  PARAM = FastMnnParam(merge.order = idx)
)
```

```{r include = FALSE}
rm(all); gc() # delete large objects and trigger garbage collection
```

Replace rowData slot:

```{r cache = FALSE}
rowData(mnn) <- DataFrame(
  gene_id = rownames(mnn),
  gene_name = mapIds(EnsDb.Mmusculus.v79, keys = rownames(mnn), column = "GENENAME", keytype = "GENEID"),
  rotation = I(rowData(mnn)$rotation)
)
```

Store combined variance model:

```{r cache = FALSE}
rownames(dec) <- rownames(mnn)

metadata(mnn)$combineVar <- dec
```

Store highly variable genes:

```{r cache = FALSE}
metadata(mnn)$getTopHVGs <- hvg
```

## Diagnostics

Check proportion of lost variance:

```{r cache = FALSE}
var <- metadata(mnn)$merge.info$lost.var

col <- colorRampPalette(RColorBrewer::brewer.pal(n = 5, name = "Reds"))(100)

brk <- seq(0, max(var), length.out = 101)

pheatmap(var[, idx], color = col, display_numbers = TRUE, number_format = "%.2f", cluster_cols = FALSE, cluster_rows = FALSE, angle_col = 0)
```

Cluster on the corrected PCA to obtain a partitioning of the cells:

```{r cache = FALSE}
snn <- buildSNNGraph(mnn, type = "jaccard", use.dimred = "corrected")

com <- igraph::cluster_louvain(snn)
```

Tabulate number of cells from multiple batches within each cluster:

```{r cache = FALSE}
tab <- table(Cluster = com$membership, Batch = mnn$batch)

col <- colorRampPalette(RColorBrewer::brewer.pal(n = 5, name = "Blues"))(100)

brk <- seq(0, max(tab), length.out = 101)

pheatmap(tab[, idx], color = col, display_numbers = TRUE, number_format = "%d", cluster_cols = FALSE, cluster_rows = FALSE, angle_col = 0)
```

Perform TSNE on the corrected PCA matrix:

```{r cache = FALSE}
set.seed(42)

mnn <- runTSNE(mnn, dimred = "corrected", perplexity = 120) # same settings as reference analysis
```

Visualize the corrected PCA using a TSNE plot:

```{r cache = FALSE, fig.width = 9, fig.height = 9}
ggcells(mnn, aes(TSNE.1, TSNE.2, colour = celltype)) + 
  geom_point(size = 0.1, show.legend = FALSE) + 
  scale_colour_manual(values = EmbryoCelltypeColours) + 
  labs(x = "UMAP 1", y = "UMAP 2") + 
  facet_wrap(~ sample) + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

Perform UMAP on the corrected PCA matrix:

```{r cache = FALSE}
set.seed(42)

mnn <- runUMAP(mnn, dimred = "corrected", n_neighbors = 20, min_dist = 0.7) # same settings as reference analysis
```

Visualize the corrected PCA using a UMAP plot:

```{r cache = FALSE, fig.width = 9, fig.height = 9}
ggcells(mnn, aes(UMAP.1, UMAP.2, colour = celltype)) + 
  geom_point(size = 0.1, show.legend = FALSE) + 
  scale_colour_manual(values = EmbryoCelltypeColours) + 
  labs(x = "UMAP 1", y = "UMAP 2") + 
  facet_wrap(~ sample) + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

## Summary

### Output

Save integrated experiment object to disk:

```{r cache = FALSE}
saveRDS(mnn, file = "output/10-atlas-integration.rds")
```

### Session

Print session information:

```{r}
sessionInfo()
```
